Fast Mini-Batch MCMC for Big Data Parameter Optimization

### 2016-05-13 nips_2016.tex
Appendix A on Gaussian Mixtures

\small
\bibliography{nips_2016}
\bibliographystyle{ieeetr}

@@

We borrow this example from~\cite{langevin_2011}. Our parameter is a two dimensional vector, denoted
as $\theta = (\theta_1, \theta_2)$, where we have
\begin{equation}
\theta_1 \sim \mathcal{N}(0, \sigma_1^2) \quad \mbox{ and } \quad \theta_2 \sim \mathcal{N}(0, \sigma_2^2)
\end{equation}
where $\mathcal{N}$ indicates the normal distribution (more generally, the multivariate normal). We
consider the above as our prior. Following~\cite{langevin_2011}, we set $\sigma_1^2 = 10$ and
$\sigma_2^2=1$, so the covariance matrix of $\theta$ is $\Sigma = {\rm diag}(10,1)$. Therefore, the
log prior probability we endow on $\theta$ is
\begin{equation}
\log p(\theta) = \log \left(\frac{1}{2\pi\sqrt{10}}\right) - \frac{1}{2}\theta^T\Sigma^{-1}\theta.
\end{equation}
To generate the data, we use the following Gaussian mixture with tied means:
\begin{equation}\label{eq:x_points}
x_i \sim \frac{1}{2}\mathcal{N}(\theta_1, \sigma_x^2) + \frac{1}{2}\mathcal{N}(\theta_1+\theta_2, \sigma_x^2)
\end{equation}
where, again following~\cite{langevin_2011}, we set $\sigma_x^2 = 2$. This means the log likelihood
of a single data instance is
\begin{equation}
\log p(x_i \mid \theta) = \log\left(\frac{1}{4\sqrt{\pi}}\right) +
\log\left(\exp\left(-\frac{1}{4}(x_i - \theta_1)^2\right) + \exp\left(-\frac{1}{4}(x_i -
(\theta_1+\theta_2))^2\right)\right)
\end{equation}
Here is the problem statement. Given some number of i.i.d. data points $x_1, x_2, \ldots, x_N$
generated according to~(\ref{eq:x_points}), determine the posterior distribution\footnote{Note that,
as is typical with Bayesian analysis, we do not concern ourselves with the denominator of the
original (non-log) posterior.} of $\theta$:
\begin{equation}\label{eq:log_post}
\log p(\theta \mid x_1,\ldots,x_n) = \log p(\theta) + \sum_{i=1}^N\log p(x_i \mid \theta).
\end{equation}
Alternatively, if there are too many data points, we may opt to instead pick a point estimate of
$\theta$, generally the MAP estimate. (If $N$ is extremely large, it will cause the posterior to
peak sharply at its modes, reducing distribution estimates to point estimates.) Note that in many
cases, we will need to take a \emph{minibatch estimate} of~(\ref{eq:log_post}). In that case, the
literature generally uses
\begin{equation}\label{eq:scaling_factor}
\log p(\theta \mid x_1,\ldots,x_n) \approx \log p(\theta) + \frac{N}{n} \sum_{i=1}^n\log p(x_i \mid \theta).
\end{equation}
where we only use $n \ll N$ samples, but we must scale up the likelihood contribution by $N/n$. If we
didn't add this scaling factor, then the contribution of the likelihood terms would be weaker. We
we will discuss this scaling factor in more detail later.

Finally, recall earlier that we needed to define $\Delta$. In our context, $\Delta$ is:
\begin{align}
\Delta &= \log \left(\frac{f(\theta') q(\theta_t \mid \theta')}{f(\theta_t) q(\theta'\mid \theta_t)} \right) \\
&= \log p(\theta') - \log p(\theta_t) + \sum_{i=1}^N(\log p(x_i \mid \theta') - \log p(x_i \mid \theta_t)) + \log\left(\frac{q(\theta_t \mid \theta')}{q(\theta' \mid \theta_t)}\right).
\end{align}

@@

Figure~\ref{fig:contour} describes a contour plot of the posterior based on the points we have
sampled. The goal is to get distributions that either cover this posterior well (i.e., identifying
that there are two modes), or if the data is too large and we want MAP estimates, we wish to see the
methods stay at one of the modes.

In the following, we only consider random walk proposals, so the term in $\Delta$ containing the
$q$s disappears.  Random walk proposals are bad, but since the proposal is poor, good performance
can only be obtained with a strong MH test, hence why we can use this as a reasonable starting
benchmark.

We use the settings:

\begin{itemize}[noitemsep]
    \item We use $N=10000$ samples of $x_i$.
    \item The covariance of the Gaussian random walk is ${\rm diag}(0.15, 0.15)$.
    \item The mini-batch size is ...
    \item (more settings) ...
    \item For the adaptive sampling method, the tolerance for deciding on a test is ....
    \item The number of iterations is .... One iteration involves one $\theta'$ proposal and a test
    to accept/reject.
\end{itemize}

Figure~\ref{fig:scatter} compares scatter plots of the ``Cutting the MH'' test (hereafter called
``adaptive minibatch'') versus ``Our Method''.

Figure~\ref{fig:cutting_mh} describes the histograms of the adaptive minibatch's batch sizes.

Figure~\ref{fig:diagnostics} describes the histograms of various statistics for our method.

\begin{figure}[ht]
  \centering
  \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
  %\includegraphics[width=0.5\linewidth]{empty}
  \caption{TODO contour plots}
  \label{fig:contour}
\end{figure}

\begin{figure}[ht]
  \centering
  \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
  %\includegraphics[width=0.5\linewidth]{empty}
  \caption{Scatter plots}
  \label{fig:scatter}
\end{figure}

\begin{figure}[ht]
  \centering
  \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
  %\includegraphics[width=0.5\linewidth]{empty}
  \caption{Diagnostic Plots 1 (e.g., histigrams of cutting MH)}
  \label{fig:cutting_mh}
\end{figure}

\begin{figure}[ht]
  \centering
  \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
  %\includegraphics[width=0.5\linewidth]{empty}
  \caption{Diagnostic Plots 2 (sd, delta, etc)}
  \label{fig:diagnostics}
\end{figure}

{\color{blue}
Daniel: TODO FINISH
}

>>>>

\small
\bibliography{nips_2016}
\bibliographystyle{ieeetr}
\normalsize

@@

We borrow this example from~\cite{langevin_2011}. Our parameter is a 2-D vector $\theta =
(\theta_1, \theta_2)$, where
\begin{equation}
\theta_1 \sim \mathcal{N}(0, \sigma_1^2) \quad \mbox{ and } \quad \theta_2 \sim \mathcal{N}(0, \sigma_2^2)
\end{equation}
where $\mathcal{N}$ indicates the normal distribution (more generally, the multivariate normal). We
consider the above as our prior. Following~\cite{langevin_2011}, we set $\sigma_1^2 = 10$ and
$\sigma_2^2=1$, so the covariance matrix of $\theta$ is $\Sigma = {\rm diag}(10,1)$. Therefore, the
log prior probability we endow on $\theta$ is
\begin{equation}
\log p(\theta) = \log \left(\frac{1}{2\pi\sqrt{10}}\right) - \frac{1}{2}\theta^T\Sigma^{-1}\theta.
\end{equation}
To generate the data, we use the following Gaussian mixture with tied means:
\begin{equation}\label{eq:x_points}
x_i \sim \frac{1}{2}\mathcal{N}(\theta_1, \sigma_x^2) + \frac{1}{2}\mathcal{N}(\theta_1+\theta_2, \sigma_x^2)
\end{equation}
where, again following~\cite{langevin_2011}, we set $\sigma_x^2 = 2$. This means the log likelihood
of a single data instance is
\begin{equation}
\log p(x_i \mid \theta) = \log\left(\frac{1}{4\sqrt{\pi}}\right) +
\log\left(\exp\left(-\frac{1}{4}(x_i - \theta_1)^2\right) + \exp\left(-\frac{1}{4}(x_i -
(\theta_1+\theta_2))^2\right)\right)
\end{equation}
Here is the problem statement. Given some number of i.i.d. data points $x_1, x_2, \ldots, x_N$
generated according to~(\ref{eq:x_points}), determine the posterior distribution\footnote{Note that,
as is typical with Bayesian analysis, we do not concern ourselves with the denominator of the
original (non-log) posterior.} of $\theta$:
\begin{equation}\label{eq:log_post}
\log p(\theta \mid x_1,\ldots,x_N) = \log p(\theta) + \sum_{i=1}^N\log p(x_i \mid \theta).
\end{equation}
Alternatively, if there are too many data points, we may opt to instead pick a point estimate of
$\theta$, generally the MAP estimate. (If $N$ is extremely large, it will cause the posterior to
peak sharply at its modes, reducing distribution estimates to point estimates.) Note that in many
cases, we will need to take a \emph{minibatch estimate} of~(\ref{eq:log_post}). In that case, the
literature generally uses
\begin{equation}\label{eq:scaling_factor}
\log p(\theta \mid x_1,\ldots,x_N) \approx \log p(\theta) + \frac{N}{n} \sum_{i=1}^n\log p(x_i \mid \theta).
\end{equation}
where we only use $n \ll N$ samples, but we must scale up the likelihood contribution by $N/n$. If we
didn't add this scaling factor, then the contribution of the likelihood terms would be weaker. There
is another thing we need to consider about that, which is the \emph{temperature} of our
distribution. In general, we will want to add $T > 1$ so that our posterior is
$p(\theta)((\prod_{i=1}^n p(x_i\mid \theta))^{N/n})^{1/T}$, resulting in the log
posterior of 
\begin{equation}\label{eq:log_prior_temp}
\log p(\theta \mid x_1,\ldots,x_N) \approx \log p(\theta) + \frac{1}{T}\frac{N}{n} \sum_{i=1}^n\log p(x_i \mid \theta).
\end{equation}
which has the extra $1/T$ to decrease the scale factor.  Equation~(\ref{eq:log_prior_temp}) is what
we will be using for our experiments, because we need to consider warmer distributions.
{\color{blue} Daniel: not totally confident on this but I'll try.}

Finally, recall earlier that we needed to define $\Delta$. In our context, $\Delta$ is:
\begin{align}
\Delta &= \log \left(\frac{f(\theta') q(\theta_t \mid \theta')}{f(\theta_t) q(\theta'\mid \theta_t)} \right) \\
&= \log p(\theta') - \log p(\theta_t) + \sum_{i=1}^N(\log p(x_i \mid \theta') - \log p(x_i \mid \theta_t)) + \log\left(\frac{q(\theta_t \mid \theta')}{q(\theta' \mid \theta_t)}\right).
\end{align}
But again, with \emph{minibatches} of data, we usually do not sum over all $N$ terms and instead sum
over $n$ terms, but with the extra $(1/T)(N/n)$ scaling factor, from
Equation~(\ref{eq:log_prior_temp}). We generally denote that approximation as $\Delta'$ in the paper.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=1\linewidth]{contour_v1}
  \caption{The posterior distribution, from 50 to 100k samples, with temperature set at 1.}
  \label{fig:contour1}
\end{figure}
\begin{figure}[t]
  \centering
  \includegraphics[width=1\linewidth]{contour_v2}
  \caption{The posterior distribution, with $N=10000$ but with temperature $T$ varying.}
  \label{fig:contour2}
\end{figure}

In this experiment, we compare our method with the one from~\cite{cutting_mh_2014}, hereafter called
the ``adaptive sampling approach''. We apply them to the Gaussian Mixture model previously
discussed.

Figure~\ref{fig:contour1} shows simulated contour plots of the posterior based on the $N$ data
points, for varying $N$, with the temperature set at $T=1$. Note that because we're using all $N$
points here, the scale factor $N/n=1$. As $N$ increases, the posterior converges to a multimodal
distribution with modes at $(0,1)$ and $(1,-1)$.

Figure~\ref{fig:contour2} is similar, except this time we fix the number of samples at $N=10000$,
but show how changing the temperature $T$ affects the distribution. A larger $T$ implies a flatter
posterior, one that (weakly) peaks in between the two true modes.

Our ultimate goal is to either (a) get a posterior distribution that covers the posterior well, or
(b) to get samples that stay trapped at a mode, for a point estimate of $\theta$.  Whether our goal
is (a) or (b) depends on our objectives and the context of the problem (e.g., big data usually
implies (b) is OK even if it would not satisfy the average Bayesian).

For both our method and the adaptive sampling method, we use random walk proposals. (Thus, the term
in $\Delta$ containing the $q$s disappears.)  Random walk proposals are bad, but since the proposal
is poor, good performance can only be obtained with a strong MH test, hence why we can use this as a
reasonable starting benchmark.

We test use the following settings:

\begin{itemize}[noitemsep]
    \item We use $N=10000$ samples of $x_i$.
    \item The covariance of the Gaussian random walk is ${\rm diag}(0.1, 0.1)$.
    \item We use 10000 iterations. One iteration involves one $\theta'$ proposal and a test to
    accept/reject.
    \item The mini-batch size is set at 100, so it is one percent of the total data. For the
    adaptive sampling approach, this is the amount that the minibatch size would keep increasing if
    we needed more precision. Our approach, of courses, keeps the size fixed.
    \item For adaptive sampling, the tolerance for deciding on a test is 0.05. The temperature for
    that is also set at $T=1$ because that algorithm was not designed to deal with temperatures.
    \item For our approach, to get an estimate of the standard deviation of $\Delta'$ (remember,
    $\Delta'$ is a minibatch approximation of $\Delta$) we take five other minibatches (randomly
    selected from the data) and take the standard deviation of those values.
\end{itemize}

For our approach, we run using three different temperature values, $T=1$, $T=10$, and $T=100$.

Figure~\ref{fig:scatter} compares scatter plots of the ``Cutting the MH'' test (hereafter called
``adaptive minibatch'') versus ``Our Method'', where we use three different temperature settings.
There doesn't seem to be too much noticeable difference in our method for the three different
temperature setting values, but our approach is a lot different from the adaptive sampling method
(this is because of rejection rates, more on that later).

\begin{figure}[ht]
  \centering
  \includegraphics[width=1\linewidth]{scatter_v01}
  \caption{Scatter plots of accepted $\theta$ for (in order) the adaptive sampling approach, and our
  method (temperatures 1, 10, and 100).}
  \label{fig:scatter}
\end{figure}

Before taking a look at our approach, we first examine the adaptive sampling approach. \textbf{That
approach rejected a lot of samples; out of 10000 iterations, it only accepted 773 times}.
Figure~\ref{fig:mb_sizes} describes the histograms of the adaptive minibatch's batch sizes. (Note
that the histogram y-axis values are different.) In many cases, using the initial batch size of 100
sufficed to make an accept/reject decision, but sometimes, this method needed to use the entire 10k
data set. Decreasing the tolerance (from its current 0.05 value) would have required larger
minibatches. Incidentally, it makes sense that the method rejects a lot of samples, because the
random walk proposal usually provides us with poor points.

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.75\linewidth]{adaptive_sampling_sizes_v01.png}
  \caption{Minibatch sizes for adaptive sampling when it was accepting or rejecting.}
  \label{fig:mb_sizes}
\end{figure}

Now we analyze our approach in more detail. Our method accepts far more samples than the adaptive
sampling method, \textbf{accepting 3285, 3318, and 3332 out of 10k}, for temperature values 1, 10,
and 100, respectively.

For diagnostics purposes, Figures~\ref{fig:diagnostics1},~\ref{fig:diagnostics2},
and~\ref{fig:diagnostics3} express histograms (for all three temperature runs) of the $\Delta'$,
${\rm std}(\Delta')$, and $X_{\rm corr}$ values, respectively.

There are a few immediate observations. First, increasing the temperature size by a factor of 10
will decrease $\Delta'$ by a factor of 10, which makes sense. That will also directly affect the
standard deviation estimate of $\Delta'$. Our $X_{\rm corr}$ distribution is symmetric about zero,
which is also expected (and it also exhibits a decrease in a factor of 10 for an increase
in temperature).

{\color{blue}
One concern I have is that our $\Delta'$ values do not seem to be ``sufficiently negative''. If our
$\Delta'$ values are negative (but large in absolute value) that means our MH test will reject more
often. But then this raises another concern: didn't we say at one point that we \emph{should} be
getting high acceptance rates, around 50 percent (in part because of the shape of the logistic
function)? Doesn't that depend on the quality of the proposal distribution? If our method is
designed to accept a lot regardless of the proposal distribution, I don't see how its performance
can be considered equal to the adaptive sampling approach, or even a standard MH test. In fact, the
best observation I can derive from this is that our approach and the adaptive sampling (or any
standard MH test) will accept the high likelihood cases, but our case will accept the borderline
ones more often.
}

\begin{figure}[ht]
  \centering
  \includegraphics[width=1\linewidth]{our_deltas_v01.png}
  \caption{Our $\Delta'$ values for three different temperature values.}
  \label{fig:diagnostics1}
\end{figure}

\begin{figure}[ht]
  \centering
  \includegraphics[width=1\linewidth]{our_sds_v01.png}
  \caption{Our ${\rm std}(\Delta')$ values for three different temperature values.}
  \label{fig:diagnostics2}
\end{figure}

\begin{figure}[ht]
  \centering
  \includegraphics[width=1\linewidth]{our_xcorrs_v01.png}
  \caption{Our $X_{\rm corr}$ values for three different temperature values.}
  \label{fig:diagnostics3}
\end{figure}
